Local PCA Shows How the Effect of Population Structure Differs Along the Genome

### 2018-08-22 writeup/local_pca_paper.tex
first pass at responding to all comments

\newif\ifsubmission
% \submissiontrue
\submissionfalse
\ifsubmission
    % % stuff for submission
    \usepackage[nofiglist,nomarkers,figuresonly]{endfloat}
    \usepackage{setspace}
    % \doublespacing
    \bibliographystyle{plos2015}
    \newcommand\citet{\cite}
    \newcommand\citep{\cite}
    \renewcommand\includegraphics{\relax}
    \renewcommand{\includegraphics}[2][]{\relax}
    \renewcommand{\figurename}{Fig.}
    \newcommand{\Figure}{Fig }
    \newcommand{\Figures}{Fig }
\else
    % for the nice version
    \usepackage[square,sort,comma,numbers]{natbib} \bibpunct{(}{)}{;}{author-year}{}{,} 
    \bibliographystyle{plainnat}
    \newcommand{\Figure}{{Figure }}
    \newcommand{\Figures}{{Figures }}
    \renewcommand{\linelabel}[1]{}
\fi
% grrr
\newcommand{\citetx}[2][]{\ifsubmission#1 \cite{#2}\else\citet{#2}\fi}

@@

Population structure leads to systematic patterns in measures of mean relatedness
between individuals in large genomic datasets,
which are often discovered and visualized using dimension reduction techniques such as principal component analysis (PCA).
Mean relatedness
is an average of the relationships across locus-specific genealogical trees,
which can be strongly affected on intermediate genomic scales by linked selection and other factors,
We show how to use local principal components analysis to describe this 
meso-scale heterogeneity in patterns of relatedness,
and apply the method to genomic data from three species,
finding in each that the effect of population structure 
can vary substantially across only a few megabases.
In a global human dataset, localized heterogeneity is likely explained by polymorphic chromosomal inversions.
In a range-wide dataset of \textit{Medicago truncatula},
factors that produce heterogeneity are shared between chromosomes,
correlate with local gene density,
and may be caused by background selection or local adaptation.
In a dataset of primarily African \textit{Drosophila melanogaster}, large-scale heterogeneity across each chromosome arm
is explained by known chromosomal inversions thought to be under recent selection,
and after removing samples carrying inversions, remaining heterogeneity is correlated with recombination rate and gene density,
again suggesting a role for linked selection.
% The method is implemented as an R package, called lostruct.
The visualization method provides a flexible new way to discover biological drivers of genetic variation,
and its application to data highlights the strong effects that linked selection and chromosomal inversions
can have on observed patterns of genetic variation.

@@

\citetx[Wright]{wright1949genetical} defined \emph{population structure} to encompass ``such matters as numbers, composition by age and sex, and state of subdivision'',
where ``subdivision'' refers to restricted migration between subpopulations.
The phrase is also commonly used to refer to the genetic patterns that result from this process,
as for instance reduced mean relatedness between individuals from distinct populations.
However, it is not necessarily clear what aspects of demography should be included in the concept.
For instance, 
\citetx[Blair]{blair1943population} defines \emph{population structure} to be the sum total of
``such factors as size of breeding populations, periodic fluctuation of population size, sex ratio, activity range and \emph{differential survival of progeny}'' (emphasis added).
The definition is similar to Wright's, but differs in including the effects of natural selection.
On closer examination, incorporating differential survival or fecundity makes the concept less clear:
should a randomly mating population consisting of two types that are partially reproductively isolated from each other
be said to show population structure or not?
Whatever the definition,
it is clear that due to natural selection, the effects of population structure --
the \emph{realized} patterns of genetic relatedness --
differ depending on which portion of the genome is being considered.
For instance, strongly locally adapted alleles of a gene will be selected against in migrants to different habitats,
increasing genetic differentiation between populations near to this gene.
Similarly, newly adaptive alleles spread first in local populations.
These observations motivate many methods to search for genetic loci under selection,
as for example in \citet{huertasnchez2013genetic}, \citet{martin2016natural}, 
and \citet{duforetfrebourg2015detecting}.

@@

These realized patterns of genetic relatedness
summarize the shapes of the genealogical trees % that relate the samples
at each location along the genome.
Since these trees vary along the genome, so does relatedness,
but averaging over sufficiently many trees we hope to get a stable estimate
that doesn't depend much on the genetic markers chosen.
This is not guaranteed:
for instance,
relatedness on sex chromosomes is expected to differ from the autosomes;
and positive or negative selection on particular loci can dramatically disort shapes of nearby genealogies
\citep{kim2002hitchhiking,charlesworth1993effect,barton2000genetic}.
Indeed,
many species show chromosome-scale variation in diversity and divergence
(e.g., \citep{langley2012genomic});
species phylogenies can differ along the genome 
due to incomplete lineage sorting,
adaptive introgression and/or local adaptation 
(e.g., \citet{pease2013accurate,ellegren2012genomic,nadeau2012genomic,pool2015natural,vernot2014resurrecting});
and theoretical expectations predict that geographic patterns of relatedness should depend on selection
\citep{charlesworth2003review}.

@@

PCA has been applied to genomic windows in methods to infer
tracts of local ancestry in recently admixed populations
\citep{bryc2010genomewide,brisbin2012pcadmix},
and to identify putative chromosomal inversions \citep{ma2012investigation}.
\revpoint{1}{8}

@@

The top 2--5 principal components are generally good summaries of population structure; 
for ease of visualization we usually only use the first two (referred to as $\pcone$ and $\pctwo$),
and check that results hold using more. \revpoint{1}{7}
The above procedure can be performed on any subset of the data;
for future reference, denote by $\pcone_j$ and $\pctwo_j$
the result after applying to all SNPs in the $j^\text{th}$ window.
(Note, however, that our measure of dissimilarity between windows
does not depend on PC ordering.) \linelabel{ll:another_pc_point}

Several of the datasets we use have unbalanced representations of diverged populations,
which can have a strong impact on the results of PCA.
(The principal axes may describe variation \emph{within} an overrepresented group
rather than more significant variation between groups.)
Therefore, to check that sampling patterns do not affect our results,
we compared to a variant of PCA that gives roughly equal weight to each group of samples,
rather than to each sample.
The rationale and implementation of this method are described in Appendix \ref{apx:weighted_pca}.

@@

We think of the local effects of population structure as being summarized by \emph{relative} position of the samples
in the space defined by the top principal components.
However, 
we do not compare patterns of relatedness of different genomic regions by directly comparing the PCs,
since rotations or reflections of these imply identical patterns of relatedness.
Instead, we compare the low-dimensional approximations of the local covariance matrices
obtained using the top $k$ PCs,
which is invariant under ordering of the PCs \revpoint{1}{3},
reflections, and rotations and yet contains all other information about the PCs.
(For results shown here, we use $k=2$;
results using larger numbers of PCs were nearly identical.)
Furthermore, to remove the effect of artifacts such as mutation rate variation,
we also rescale each approximate covariance matrix to be of similar size
(precisely, so that the underlying data matrix has trace norm equal to one).

To do this, define the $N \times k$ matrix $V(i)$ so that $V(i)_{\cdot \ell}$, 
the $\ell^\text{th}$ column of $V(i)$,
is equal to the $\ell^\text{th}$ princpal component of the $i^\text{th}$ window,
multiplied by $( \lambda_{\ell i} / \sum_{m=1}^k \lambda_{m i} )^{1/2}$,
where $\lambda_{\ell i}$ is the $\ell^\text{th}$ eigenvalue of the genetic covariance matrix.
% \begin{equation}
%     \begin{aligned}
%         V_{i1}&=\sqrt{\frac{\lambda _{1i}}{\lambda _{1i}+\lambda _{2i}}}PC1_{i} ,
%         \qquad
%         V_{i2}&=\sqrt{\frac{\lambda _{2i}}{\lambda _{1i}+\lambda _{2i}}}PC2_{i} 
%     \end{aligned}
% \end{equation}
Then, the rescaled, rank $k$ approximate covariance matrix for the $i^\text{th}$ window is
\begin{align} \label{eqn:est_cov}
    M(i) &= \sum_{\ell=1}^k V(i)_{\cdot \ell} V(i)_{\cdot \ell}^T .
\end{align}
% \begin{align}
%     M_{i} &= \frac{\lambda_{1i}PC1_{i}PC1_{i}^{T}+\lambda_{2i}PC2_{i}PC2_{i}^{T}}{\lambda_{1i}+\lambda_{2i}} 
% \end{align}

To measure the similarity of patterns of relatedness for the $i^\text{th}$ window and $j^\text{th}$ window,
we then use
Euclidean distance $D_{ij}$ between the matrices $M(i)$ and $M(j)$:
$D_{ij}^2 = \sum_{k\ell} ( M(i)_{k,\ell} - M(j)_{k,\ell} )^2$.

@@

%%%%%%%%%%%%%%%%%%%
\subsection{Testing}

@@

To provide a realistic test,
we next used forwards-time, individual-based simulations, implemented using SLiM v3 \citep{slim_v3},
which are described in detail in Appendix \ref{apx:slim_sims}.
To provide realistic population structure for PCA to identify,
each simulation ahd at least 5,000 diploid individuals,
living across a continuous square range,
with Gaussian dispersal and local density-dependent competition.
Each genome was modeled on human chromosome 7, which is $1.54 \times 10^8$bp long,
with an overall recombination rate of 1.6785 crossovers per chromosome per generation.
To improve speed, we used tskit \citep{tskit_paper} to record tree sequences
and add neutral mutations afterwards, at a rate of $10^{-9}$ per bp per generation.
Most simulations were neutral, but we also included linked selection, of two types.
First, % run_031486
we introduced selected mutations into two regions, 
which extended from 1/3 to 1/2 and from 5/6 to the end of the genome respectively.
These had selection coefficients from a Gamma distribution with shape 2 and mean 0.005
at a rate of $10^{-10}$ per bp, that were either beneficial (with probability $1/30$)
or deleterious (otherwise).
Second, % run_016913
to roughly model a recent expansion followed by local adaptation,
we introduced mutations in the same manner as above,
except that mutations were no longer unconditionally deleterious or beneficial:
each selection coefficient was multiplied by a factor depending on the spatial location
of the individual being evaluated, varying linearly from -1 at the left side of the range
to +1 at the right edge.
In all simulations, genome-wide PCA displayed a map of the population range, as expected.

@@

%%%%%%%%%%%
\subsection{Validation}

\plr{Address:
    mutation rate variation;
    recombination rate variation;
    choice of number of PCs;
    choice of window size;
    variation in missingness;
    maybe differing sample sizes and reweighting.  }

Simple non-population-based simulations with Gaussian ``genotypes''
showed that the method performs as expected, clearly separating regions of the genome
with different underlying covariance matrices
without being affected by extreme differences in amount of missing data
(Supplemental Figure \ref{fig:test_missing}).
This simulation also verifies insensitivity to ordering of top PCs,
since it was performed using a covariance matrix with the top two eigenvalues equal,
so that the order of empirical eigenvectors (PCs) switches randomly.
The same type of simulations with dramatically different sample sizes
showed that using weighted PCA substantially improved identification of different genomic regions
(Supplemental Figure \ref{fig:test_weighted_pca}).

Individual-based simulations using SLiM \citep{slim_v3} allowed us to test the effects
of recombination and mutation rate variation, as well as linked selection.
Results are shown in Supplemental Figures \ref{sfig:slim_sims1} and \ref{sfig:slim_sims2}.
As expected, varying recombination rate stepwise by a factor of 64
did not induce patterns in the MDS visualizations correlated with recombination rate.
Since varying mutation rate with a fixed recombination map
is equivalent to varying the recombination map and remapping windows,
we used the same simulations to show that variation in mutation rate similarly has no effect.
On the other hand,
a recombination map with hotspots
(the HapMap human female map for chromosome 7 \citep{hapmap})
induced outliers at long regions of low recombination rate
(also as expected).

Simulations with linked selection produced mixed results
(Figure \ref{sfig:linked_selection}).
The method strongly identified the regions under spatially varying linked selection.
It also identified the regions (although less unambiguously)
with constant selection and stepwise varying recombination rate,
but did not clearly identify them with constant recombination rate.
This difference is likely because recombination rates are overall lower in the first case,
leading to a stronger effect of linked seleciton.
These tests are not meant to be comprehensive survey of linked selection,
but only to demonstrate that linked selection can produce signals simlar to what we see in real data.

@@

We applied the method to windows of average length 9 Kbp, across chromosome arms 2L, 2R, 3L, 3R and X separately.
The first column of \Figure \ref{fig:mds_allchr} is a multidimensional scaling (MDS) visualization 
of the matrix of dissimilarities between genomic windows:
in other words, genomic windows that are closer to each other in the MDS plot show more similar patterns of relatedness.
For each chromosome arm, the MDS visualization roughly resembles a triangle,
sometimes with additional points.
Since the relative position of each window in this plot shows the similarity between windows, 
this suggests that there are at least three extreme manifestations of population structure 
typified by windows found in the ``corners'' of the figure,
and that other windows' patterns of relatedness may be a mixture of those extremes. 
The next two columns of \Figure \ref{fig:mds_allchr} respectively depict the two MDS coordinates of each window,
plotted against the window's position along the genome,
to show how the plot of the first column is laid out along the genome.

@@

The most striking variation in patterns of relatedness turns out to be explained by
several large inversions that are polymorphic in these samples, 
discussed in \citet{corbett2012population} and \citet{langley2012genomic}.
% Frequencies of each inversion in the sample are given in Table \ref{tab:inversion_freqs}.
To depict this, \Figure \ref{fig:pca_by_inversion} shows
the PCA plots in \Figure \ref{fig:pca_by_pop} recolored by the orientation of the inversion for each sample.
Taking chromosome arm 2L as an example,
the two regions of similar, extreme patterns of relatedness
shown in green in the first row of \Figure \ref{fig:mds_allchr}
lie directly around the breakpoints of the inversion In(2L)t,
and the PCA plots in the first rows of \Figure \ref{fig:pca_by_inversion}
shows that patterns of relatedness here are mostly determined by inversion orientation.
The regions shown in purple on chromosome 2L lie near the centromere,
and have patterns of relatedness reflective of two axes of variation,
seen in \Figures \ref{fig:pca_by_pop} and \ref{fig:pca_by_inversion},
which correspond roughly to latitude within Africa and to degree of cosmopolitan admixture respectively
(see \citet{lack2015drosophila} for more about admixture in this sample).
The regions shown in orange on chromosome 2L mostly lie inside the inversion,
and show patterns of relatedness that are a mixture between the other two,
as expected due to recombination within the (long) inversion \citep{guerrero2011coalescent}.
Similar results are found in other chromosome arms,
albeit complicated by the coexistence of more than one polymorphic inversion;
however, each breakpoint visibly affects patterns in the MDS coordinates
(see vertical lines in \Figure \ref{fig:mds_allchr}).
% \plr{\citet{corbett2012population} says In(1)Be is very young, $\sim$ 60 years.
% Also cite \citet{corbettdetig2012sequencebased} for a computational method for detecting inversions.}
% corbettdetig2012sequencebased suggests these inversions are recent-ish (but before out-of-Africa)
% except that the situation of Chr3R is a little more complicated,
% due to the coexistence of two polymorphic inversions (In(3R)K and In(3R)P) at intermediate frequency.
% However, peaks in MDS coordinates along the chromosome show the breakpoints of both,
% although we have not highlighted portions of the MDS plot that pick these out.
% The inversions In(3L)P and In(3R)Mo, are at low frequency in these samples, 
% and visibly affect the MDS plots although less than the more common inversion on each arm.
% The very young inversion In(X)be shows little signal 
% compared to the older inversion In(X)A \citep{corbett2012population}.

@@

Unlike the other two species,
the method applied separately on all eight chromosomes of \textit{Medicago truncatula} 
showed similar patterns of gradual change in patterns of relatedness across each chromosome,
with no indications of chromosome-specific patterns.
This consistency suggests that the factor affecting the population structure for each chromosome is the same,
as might be caused by varying strengths of linked selection.
To verify that variation in the effects of population structure is shared across chromosomes,
we applied the method to all chromosomes together.
Results for chromosome 3 are shown in \Figures \ref{fig:mds12_medicago} and \ref{fig:pca_medicago},
and other chromosomes are similar:
% the two types of population structure with higher gene density 
% (colored green and purple in \Figure \ref{fig:mds12_medicago})
% are nearly identical across all chromosomes,
% but population structure associated with low gene density (colored orange)
% shows more chromsome-specific patterns (see Supplemental \Figure \ref{sfig:pca_drosophila_allchr}).
across chromosomes, the high values of the first MDS coordinate coincide with the position of the heterochromatic regions surrounding the centromere,
which often have lower gene density and may therefore be less subject to linked selection.
% Selection on genes in Medicago has been demonstrated by \citet{paape2013selection}. 
To verify that this is a possible explanation,
we counted the number of genes found in each window using gene models in Mt4.0 from \url{jcvi.org} \citep{tang2014improved},
which are shown juxtaposed with
the first MDS coordinate of each window in \Figure \ref{fig:mds_medicago},
and are significantly correlated, as shown in Supplemental \Figure \ref{sfig:mds_gene_count}.
(Values shown are the number of start and end positions of each predicted mRNA transcript,
divided by two, assigned to the nearest window.)
% that lie closer to SNPs  between the last SNP of each window and the first SNP in the next window.
However, other genomic features, such as distance to centromere show roughly the same patterns,
so we cannot rule out alternative hypotheses.
In particular, fine-scale recombination rate estimates are not available in a form mappable
to Mt4.0 coordinates (although those in \citet{paape2012finescale} appear visually similar).
\revpoint{2}{8}

We also found nearly identical results when choosing shorter windows of 1,000 SNPs;
or choosing windows of equal length in base pairs rather than SNPs.
Similarly, the results were not substantially changed
when using weighted PCA to downweight the large group of Tunisian samples.

>>>>

\newif\ifsubmission
% \submissiontrue
\submissionfalse
\ifsubmission
        \usepackage[nofiglist,nomarkers,figuresonly]{endfloat}
        \renewcommand\includegraphics{\relax}
        \renewcommand{\includegraphics}[2][]{\relax}
        \renewcommand{\figurename}{Fig.}
    \bibliographystyle{plos2015}
    \newcommand\citet{\cite}
    \newcommand\citep{\cite}
    \newcommand{\Figure}{Fig }
    \newcommand{\Figures}{Fig }
\else
    % for the nice version
    \usepackage[square,sort,comma,numbers]{natbib} 
    % \bibpunct{(}{)}{;}{author-year}{}{,} 
    \bibliographystyle{plainnat}
    \newcommand{\Figure}{{Figure }}
    \newcommand{\Figures}{{Figures }}
    % \renewcommand{\linelabel}[1]{}
\fi
% grrr
\newcommand{\citetx}[2][]{\ifsubmission#1 \cite{#2}\else\citet{#2}\fi}

@@

Population structure leads to systematic patterns in measures of mean relatedness
between individuals in large genomic datasets,
which are often discovered and visualized using dimension reduction techniques such as principal component analysis (PCA).
Mean relatedness
is an average of the relationships across locus-specific genealogical trees,
which can be strongly affected on intermediate genomic scales by linked selection and other factors.
We show how to use local principal components analysis to describe this 
meso-scale heterogeneity in patterns of relatedness,
and apply the method to genomic data from three species,
finding in each that the effect of population structure 
can vary substantially across only a few megabases.
In a global human dataset, localized heterogeneity is likely explained by polymorphic chromosomal inversions.
In a range-wide dataset of \textit{Medicago truncatula},
factors that produce heterogeneity are shared between chromosomes,
correlate with local gene density,
and may be caused by background selection or local adaptation.
In a dataset of primarily African \textit{Drosophila melanogaster}, large-scale heterogeneity across each chromosome arm
is explained by known chromosomal inversions thought to be under recent selection,
and after removing samples carrying inversions, remaining heterogeneity is correlated with recombination rate and gene density,
again suggesting a role for linked selection.
% The method is implemented as an R package, called lostruct.
The visualization method provides a flexible new way to discover biological drivers of genetic variation,
and its application to data highlights the strong effects that linked selection and chromosomal inversions
can have on observed patterns of genetic variation.

@@

\citetx[Wright]{wright1949genetical} defined \emph{population structure} to encompass ``such matters as numbers, composition by age and sex, and state of subdivision'',
where ``subdivision'' refers to restricted migration between subpopulations.
The phrase is also commonly used to refer to the genetic patterns that result from this process,
as for instance reduced mean relatedness between individuals from distinct populations.
However, it is not necessarily clear what aspects of demography should be included in the concept. \revpoint{2}{10}
For instance, 
\citetx[Blair]{blair1943population} defines \emph{population structure} to be the sum total of
``such factors as size of breeding populations, periodic fluctuation of population size, sex ratio, activity range and \emph{differential survival of progeny}'' (emphasis added).
The definition is similar to Wright's, but differs in including the effects of natural selection.
On closer examination, incorporating differential survival or fecundity makes the concept less clear:
should a randomly mating population consisting of two types that are partially reproductively isolated from each other
be said to show population structure or not?
Whatever the definition,
it is clear that due to natural selection, the effects of population structure --
the \emph{realized} patterns of genetic relatedness --
differ depending on which portion of the genome is being considered.
For instance, strongly locally adapted alleles of a gene will be selected against in migrants to different habitats,
increasing genetic differentiation between populations near to this gene.
Similarly, newly adaptive alleles spread first in local populations.
These observations motivate many methods to search for genetic loci under selection,
as for example in \citet{huertasnchez2013genetic}, \citet{martin2016natural}, 
and \citet{duforetfrebourg2015detecting}.

@@

These realized patterns of genetic relatedness
summarize the shapes of the genealogical trees % that relate the samples
at each location along the genome.
Since these trees vary along the genome, so does relatedness,
but averaging over sufficiently many trees we hope to get a stable estimate
that doesn't depend much on the genetic markers chosen.
This is not guaranteed:
for instance,
relatedness on sex chromosomes is expected to differ from the autosomes;
and positive or negative selection on particular loci can dramatically disort shapes of nearby genealogies
\citep{kim2002hitchhiking,charlesworth1993effect,barton2000genetic}.
Indeed,
many species show chromosome-scale variation in diversity and divergence
(e.g., \citet{langley2012genomic});
species phylogenies can differ along the genome 
due to incomplete lineage sorting,
adaptive introgression and/or local adaptation 
(e.g., \citet{pease2013accurate,ellegren2012genomic,nadeau2012genomic,pool2015natural,vernot2014resurrecting});
and theoretical expectations predict that geographic patterns of relatedness should depend on selection
\citep{charlesworth2003review}.

@@

Others have applied PCA to windows of the genome:
\citet{ma2012investigation} used local PCA much as we do 
to identify putative chromosomal inversions.
\citet{bryc2010genomewide} and \citet{brisbin2012pcadmix}
use PCA to infer tracts of local ancestry in recently admixed populations,
but by projecting each genomic window onto the axes of a single, globally-defined PCA
rather than doing PCA separately on each window.  \revpoint{1}{7}

@@

The top 2--5 principal components are generally good summaries of population structure; 
for ease of visualization we usually only use the first two (referred to as $\pcone$ and $\pctwo$),
and check that results hold using more. \revpoint{1}{6}
The above procedure can be performed on any subset of the data;
for future reference, denote by $\pcone_j$ and $\pctwo_j$
the result after applying to all SNPs in the $j^\text{th}$ window.
(Note, however, that our measure of dissimilarity between windows
does not depend on PC ordering.) \linelabel{ll:another_pc_point}

% Several of the datasets we use have unbalanced representations of diverged populations,
% which can have a strong impact on the results of PCA.
% (The principal axes may describe variation \emph{within} an overrepresented group
% rather than more significant variation between groups.)
% Therefore, to check that sampling patterns do not affect our results,
% we compared to a variant of PCA that gives roughly equal weight to each group of samples,
% rather than to each sample.
% The rationale and implementation of this method are described in Appendix \ref{apx:weighted_pca}.

@@

We think of the local effects of population structure as being summarized by \emph{relative} position of the samples
in the space defined by the top principal components.
However, 
we do not compare patterns of relatedness of different genomic regions by directly comparing the PCs,
since rotations or reflections of these imply identical patterns of relatedness.
Instead, we compare the low-dimensional approximations of the local covariance matrices
obtained using the top $k$ PCs,
which is invariant under ordering of the PCs \revpoint{1}{2},
reflections, and rotations and yet contains all other information about the PCs.
(For results shown here, we use $k=2$;
results using larger numbers of PCs were nearly identical.)
Furthermore, to remove the effect of artifacts such as mutation rate variation,
we also rescale each approximate covariance matrix to be of similar size
(precisely, so that the underlying data matrix has trace norm equal to one).

To do this, define the $N \times k$ matrix $V(i)$ so that $V(i)_{\cdot \ell}$, 
the $\ell^\text{th}$ column of $V(i)$,
is equal to the $\ell^\text{th}$ princpal component of the $i^\text{th}$ window,
multiplied by $( \lambda_{\ell i} / \sum_{m=1}^k \lambda_{m i} )^{1/2}$,
where $\lambda_{\ell i}$ is the $\ell^\text{th}$ eigenvalue of the genetic covariance matrix.
% \begin{equation}
%     \begin{aligned}
%         V_{i1}&=\sqrt{\frac{\lambda _{1i}}{\lambda _{1i}+\lambda _{2i}}}PC1_{i} ,
%         \qquad
%         V_{i2}&=\sqrt{\frac{\lambda _{2i}}{\lambda _{1i}+\lambda _{2i}}}PC2_{i} 
%     \end{aligned}
% \end{equation}
Then, the rescaled, rank $k$ approximate covariance matrix for the $i^\text{th}$ window is
\begin{align} \label{eqn:est_cov}
    M(i) &= \sum_{\ell=1}^k V(i)_{\cdot \ell} V(i)_{\cdot \ell}^T .
\end{align}
% \begin{align}
%     M_{i} &= \frac{\lambda_{1i}PC1_{i}PC1_{i}^{T}+\lambda_{2i}PC2_{i}PC2_{i}^{T}}{\lambda_{1i}+\lambda_{2i}} 
% \end{align}
To measure the similarity of patterns of relatedness for the $i^\text{th}$ window and $j^\text{th}$ window,
we then use
Euclidean distance $D_{ij}$ between the matrices $M(i)$ and $M(j)$:
$D_{ij}^2 = \sum_{k\ell} ( M(i)_{k,\ell} - M(j)_{k,\ell} )^2$.

@@

%%%%%%%%%%%%%%%%%%%
\subsection{Testing}
\label{ss:testing_methods}

@@

To provide a realistic test,
we next used forwards-time, individual-based simulations, implemented using SLiM v3 \citep{haller2017flexible},
which are described in detail in Appendix \ref{apx:slim_sims}.
To provide realistic population structure for PCA to identify,
each simulation ahd at least 5,000 diploid individuals,
living across a continuous square range,
with Gaussian dispersal and local density-dependent competition.
Each genome was modeled on human chromosome 7, which is $1.54 \times 10^8$bp long,
with an overall recombination rate of 1.6785 crossovers per chromosome per generation.
To improve speed, we used tskit \citep{kelleher2018efficient} to record tree sequences
and add neutral mutations afterwards, at a rate of $10^{-9}$ per bp per generation.
Most simulations were neutral, but we also included linked selection, of two types.
First, % run_031486
we introduced selected mutations into two regions, 
which extended from 1/3 to 1/2 and from 5/6 to the end of the genome respectively.
These had selection coefficients from a Gamma distribution with shape 2 and mean 0.005
at a rate of $10^{-10}$ per bp, that were either beneficial (with probability $1/30$)
or deleterious (otherwise).
Second, % run_016913
to roughly model a recent expansion followed by local adaptation,
we introduced mutations in the same manner as above,
except that mutations were no longer unconditionally deleterious or beneficial:
each selection coefficient was multiplied by a factor depending on the spatial location
of the individual being evaluated, varying linearly from -1 at the left side of the range
to +1 at the right edge.
In all simulations, genome-wide PCA displayed a map of the population range, as expected.

@@

%%%%%%%%%%%
\subsection{Validation}
\label{ss:validation}

Simple non-population-based simulations with Gaussian ``genotypes''
showed that the method performs as expected, clearly separating regions of the genome
with different underlying covariance matrices
without being affected by extreme differences in amount of missing data
(Supplemental Figure \ref{fig:test_missing}).
This simulation also verifies insensitivity to ordering of top PCs,
since it was performed using a covariance matrix with the top two eigenvalues equal,
so that the order of empirical eigenvectors (PCs) switches randomly.

Individual-based simulations using SLiM \citep{haller2017flexible} allowed us to test the effects
of recombination and mutation rate variation, as well as linked selection.
Results are shown in Supplemental Figures \ref{sfig:slim_sims1} and \ref{sfig:slim_sims2}.
As expected, varying recombination rate stepwise by a factor of 64
did not induce patterns in the MDS visualizations correlated with recombination rate.
\linelabel{ll:test_recomb_rate}
Since varying mutation rate with a fixed recombination map
is equivalent to varying the recombination map and remapping windows,
we used the same simulations to show that variation in mutation rate similarly has no effect.
On the other hand,
a recombination map with hotspots
(the HapMap human female map for chromosome 7 \citep{hapmap2007second})
induced outliers at long regions of low recombination rate
(also as expected). \linelabel{ll:hotspot_recomb}

Simulations with linked selection produced mixed results
(Figure \ref{sfig:slim_sims2}).
The method strongly identified the regions under spatially varying linked selection.
It also identified the regions (although less unambiguously)
with constant selection and stepwise varying recombination rate,
but did not clearly identify them with constant recombination rate.
This difference is likely because recombination rates are overall lower in the first case,
leading to a stronger effect of linked seleciton.
These tests are not meant to be comprehensive survey of linked selection,
but only to demonstrate that linked selection can produce signals simlar to what we see in real data.

@@

We applied the method to windows of average length 9 Kbp, across chromosome arms 2L, 2R, 3L, 3R and X separately.
The first column of \Figure \ref{fig:mds_allchr} is a multidimensional scaling (MDS) visualization 
of the matrix of dissimilarities between genomic windows:
in other words, genomic windows that are closer to each other in the MDS plot show more similar patterns of relatedness.
For each chromosome arm, the MDS visualization roughly resembles a triangle,
sometimes with additional points.
Since the relative position of each window in this plot shows the similarity between windows, 
this suggests that there are at least three extreme manifestations of population structure 
typified by windows found in the ``corners'' of the figure,
and that other windows' patterns of relatedness may be a mixture of those extremes. 
The next two columns of \Figure \ref{fig:mds_allchr} respectively depict the two MDS coordinates of each window,
plotted against the window's position along the genome,
to show how the plot of the first column is laid out along the genome.
The patterns did not depend on the number of PCs used
(see \Figure \ref{fig:dmel_k5} for the same plot with $k=5$ PCs), \revpoint{1}{3}
and are not driven by variation in missingness (see \Figure \ref{sfig:dpgp_missingness}). \revpoint{2}{4}

@@

The most striking variation in patterns of relatedness turns out to be explained by
several large inversions that are polymorphic in these samples, 
discussed in \citet{corbett2012population} and \citet{langley2012genomic}.
% Frequencies of each inversion in the sample are given in Table \ref{tab:inversion_freqs}.
To depict this, \Figure \ref{fig:pca_by_inversion} shows
the PCA plots in \Figure \ref{fig:pca_by_pop} recolored by the orientation of the inversion for each sample.
Taking chromosome arm 2L as an example,
the two regions of similar, extreme patterns of relatedness
shown in green in the first row of \Figure \ref{fig:mds_allchr}
lie directly around the breakpoints of the inversion In(2L)t,
and the PCA plots in the first rows of \Figure \ref{fig:pca_by_inversion}
shows that patterns of relatedness here are mostly determined by inversion orientation.
The regions shown in purple on chromosome 2L lie near the centromere,
and have patterns of relatedness reflective of two axes of variation,
seen in \Figures \ref{fig:pca_by_pop} and \ref{fig:pca_by_inversion},
which correspond roughly to latitude within Africa and to degree of cosmopolitan admixture respectively
(see \citet{lack2015drosophila} for more about admixture in this sample).
The regions shown in orange on chromosome 2L mostly lie inside the inversion,
and show patterns of relatedness that are a mixture between the other two,
as expected due to recombination within the (long) inversion \citep{guerrero2011coalescent}.
Similar results are found in other chromosome arms,
albeit complicated by the coexistence of more than one polymorphic inversion;
however, each breakpoint visibly affects patterns in the MDS coordinates
(see vertical lines in \Figure \ref{fig:mds_allchr}).

% \plr{\citet{corbett2012population} says In(1)Be is very young, $\sim$ 60 years.
% Also cite \citet{corbettdetig2012sequencebased} for a computational method for detecting inversions.}
% corbettdetig2012sequencebased suggests these inversions are recent-ish (but before out-of-Africa)
% except that the situation of Chr3R is a little more complicated,
% due to the coexistence of two polymorphic inversions (In(3R)K and In(3R)P) at intermediate frequency.
% However, peaks in MDS coordinates along the chromosome show the breakpoints of both,
% although we have not highlighted portions of the MDS plot that pick these out.
% The inversions In(3L)P and In(3R)Mo, are at low frequency in these samples, 
% and visibly affect the MDS plots although less than the more common inversion on each arm.
% The very young inversion In(X)be shows little signal 
% compared to the older inversion In(X)A \citep{corbett2012population}.

@@

Unlike the other two species,
the method applied separately on all eight chromosomes of \textit{Medicago truncatula} 
showed similar patterns of gradual change in patterns of relatedness across each chromosome,
with no indications of chromosome-specific patterns.
This consistency suggests that the factor affecting the population structure for each chromosome is the same,
as might be caused by varying strengths of linked selection.
To verify that variation in the effects of population structure is shared across chromosomes,
we applied the method to all chromosomes together.
Results for chromosome 3 are shown in \Figure \ref{fig:mds12_medicago}, \revpoint{2}{13}
and other chromosomes are similar:
across chromosomes, the high values of the first MDS coordinate coincide with the position of the heterochromatic regions surrounding the centromere,
which often have lower gene density and may therefore be less subject to linked selection.
% Selection on genes in Medicago has been demonstrated by \citet{paape2013selection}. 
To verify that this is a possible explanation,
we counted the number of genes found in each window using gene models in Mt4.0 from \url{jcvi.org} \citep{tang2014improved},
which are shown juxtaposed with
the first MDS coordinate of each window in \Figure \ref{fig:mds_medicago},
and are significantly correlated, as shown in Supplemental \Figure \ref{sfig:mds_gene_count}.
(Values shown are the number of start and end positions of each predicted mRNA transcript,
divided by two, assigned to the nearest window.)
% that lie closer to SNPs  between the last SNP of each window and the first SNP in the next window.
However, other genomic features, such as distance to centromere show roughly the same patterns,
so we cannot rule out alternative hypotheses.
In particular, fine-scale recombination rate estimates are not available in a form mappable
to Mt4.0 coordinates (although those in \citet{paape2012finescale} appear visually similar).
\revpoint{2}{8}

We also found nearly identical results when choosing shorter windows of 1,000 SNPs
(Supplemental Figure \ref{sfig:mds_medicago_allchr_1e3}; \revpoint{1}{4}
or choosing windows of equal length in base pairs rather than SNPs. \revpoint{1}{5}
% Similarly, the results were not substantially changed
% when using weighted PCA to downweight the large group of Tunisian samples.

\begin{figure}
    \begin{center}
       \includegraphics{Fig6_Together_MDS_plot_chr3_final}
       \includegraphics{Fig7_pca_plots_for_Medicago_chr3_3peaks_byMDS}
    \end{center}
    \caption{
        \textbf{(top)}
       MDS visualization of patterns of relatedness on \textit{M.~truncatula} chromosome 3,
       with corresponding PCA plots.
       Each point in the plot represents a window;
       the structure revealed by the MDS plot is strongly clustered along the chromosome,
       with windows in the upper-right corner of the MDS plot (colored red) clustered around the centromere,
       windows in the upper-left corner (purple) furthest from the centromere,
       and the remaining corner (green) intermediate.
       Plots for remaining chromosomes are shown in Supplemental \Figure \ref{sfig:mds_medicago_allchr}.
       \textbf{(bottom)}
        PCA plots for the sets of genomic windows colored (A) green, (B) orange, and (C) purple in the top figure.
        Each point corresponds to a sample, colored by country of origin.
        Plots for remaining chromosomes are shown in Supplemental \Figure \ref{sfig:pca_peaks_medicago_allchr}.
       \label{fig:mds12_medicago}
       \label{fig:pca_medicago}
    }
\end{figure}
